---
title: "The ggseg_atlas object"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The ggseg_atlas object}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup-opts
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r}
#| label: setup
library(ggseg.formats)
# sf is a Suggests dependency since the sf-optional milestone. Only the two
# chunks below that render through the sf path need it, so they are skipped
# when it is unavailable rather than failing the whole vignette.
has_sf <- requireNamespace("sf", quietly = TRUE)
```

Everything in the ggsegverse ecosystem starts from a single object: the `ggseg_atlas`.
Whether you are making a 2D cortical flatmap with ggseg or spinning a 3D mesh in
ggseg3d, the atlas is the container that holds the geometry, the region metadata,
and the colour palette together. This vignette walks through the structure so you
know exactly what you are working with.

## What is a ggseg_atlas?

A `ggseg_atlas` is an S3 object that bundles five pieces of information into one
handle. Let's print the bundled Desikan-Killiany atlas to see what that looks
like:

```{r}
#| label: print-dk
dk()
```
The print method gives you a quick overview: the atlas name, type, how many
regions it has, which hemispheres are present, what views the 2D geometry
provides, and whether palette and rendering data are available. Below the summary
you see the core table, the single source of truth for region identity.

## Anatomy of an atlas

This section reaches into the five slots with `$` on purpose: its subject is
the object layout, and `$` is the only way to show it. Everywhere else, use the
accessors in [Querying an atlas](#querying-an-atlas) — they are the supported
interface and keep working when the layout changes.

```{r}
#| label: atlas-slots
dk()$atlas
dk()$type
```

`$atlas` is a short name (used to look up atlases by string) and `$type` is one
of the values `atlas_types()` returns: `"cortical"`, `"subcortical"`,
`"tract"`, or `"cerebellar"`.

The `$palette` is a named character vector mapping labels to hex colours:

```{r}
#| label: palette
head(dk()$palette)
```

`$core` is a data frame with one row per region. It always has `region` and
`label` columns, and will often include `hemi` and additional metadata like
`lobe` or `structure`:

```{r}
#| label: core
head(dk()$core)
```

Finally, `$data` is a `ggseg_atlas_data` object that holds the actual geometry.
Its contents depend on the atlas type.

```{r}
#| label: data-class
class(dk()$data)
```

## Three atlas types

ggseg.formats ships three atlases that illustrate the three types.

**Cortical** atlases like `dk` parcellate the cortical surface. Their data object
is a `ggseg_data_cortical` containing sf polygons for 2D rendering and vertex indices
for 3D:

```{r}
#| label: cortical-type
dk()$type
names(dk()$data)
```

**Subcortical** atlases like `aseg` represent deep brain structures. Their data
is a `ggseg_data_subcortical` with sf polygons and individual 3D meshes:

```{r}
#| label: subcortical-type
aseg()$type
names(aseg()$data)
```

**Tract** atlases like `tracula` represent white matter bundles. Their data is a
`ggseg_data_tract` with sf polygons and centerlines that generate tube meshes for 3D:

```{r}
#| label: tract-type
tracula()$type
names(tracula()$data)
```

In every case the sf component drives 2D plotting and the type-specific component
(vertices, meshes, or centerlines) drives 3D.

## Core: the region table

The `$core` data frame is the single source of truth for what regions an atlas
contains. Every manipulation function updates core first and then propagates
changes to geometry and palette.

The required columns are `region` (a short, hemisphere-free key), `label` (a
unique identifier that links core to geometry and palette) and `names` (the
curated long-form display name). Most atlases also carry `hemi`:

```{r}
#| label: core-structure
str(dk()$core)
```

Some atlases include additional metadata columns. The `dk` atlas, for instance,
has `lobe`:

```{r}
#| label: core-lobe
unique(dk()$core$lobe)
```

You can add your own metadata with `atlas_core_add()` (covered in the
manipulation vignette).

## Querying an atlas

A set of accessor functions lets you pull information out. This, not the `$`
access shown above, is the supported way to read an atlas: it survives changes
to the internal layout, and it is what the rest of the ggsegverse uses.

Each of these returns one element per `core` row, in `core` row order, so they
are row-aligned with each other: `atlas_labels(a)[i]`, `atlas_regions(a)[i]` and
`atlas_names(a)[i]` all describe the same row. Wrap a call in `unique()` when
you want the distinct set instead.

`atlas_regions()` returns the region keys:

```{r}
#| label: atlas-regions
unique(atlas_regions(dk()))
```

`atlas_labels()` returns the labels (the identifiers used to join geometry):

```{r}
#| label: atlas-labels
head(atlas_labels(dk()))
```

`atlas_names()` returns the curated long-form display names — what you want for
a figure legend:

```{r}
#| label: atlas-names
head(atlas_names(aseg()))
```

`region` held long-form names itself until version 0.1.0. `legacy_region_map()`
translates those old values to the current keys, which is what you need to fix a
table or script written against an earlier release. It reads a recorded table of
what each bundled atlas shipped at 0.0.4, so it is not the same thing as
`names`:

```{r}
#| label: legacy-region-map
head(legacy_region_map(aseg()))
```

`atlas_views()` returns the available 2D views:

```{r}
#| label: atlas-views
atlas_views(dk())
atlas_views(aseg())
atlas_views(tracula())
```

`atlas_type()` returns the type string:

```{r}
#| label: atlas-type
atlas_type(dk())
atlas_type(aseg())
atlas_type(tracula())
```

`atlas_palette()` retrieves the colour palette from an atlas object:

```{r}
#| label: atlas-palette
head(atlas_palette(dk()))
```

## Extracting render-ready data

When you need the actual data frames that ggseg and ggseg3d consume, use the
`atlas_*()` extractors. These join core metadata and palette colours onto the raw
geometry so you get a single, ready-to-use table.

`atlas_sf()` returns an sf data frame for 2D rendering. It is the one extractor
that requires sf; `atlas_polygons()` is its sf-free counterpart and returns the
same geometry in the `brain_polygons` representation the bundled atlases ship in:

```{r}
#| label: atlas-sf
#| eval: !expr has_sf
sf_data <- atlas_sf(dk())
sf_data
```

`atlas_vertices()` returns the vertex data for cortical 3D rendering:

```{r}
#| label: atlas-vertices
vert_data <- atlas_vertices(dk())
vert_data
```

`atlas_meshes()` returns mesh data for subcortical or tract 3D rendering:

```{r}
#| label: atlas-meshes
mesh_data <- atlas_meshes(aseg())
mesh_data
```

`as.data.frame()` is a convenience method that produces a merged sf data frame
similar to `atlas_sf()` but with atlas-level columns (`atlas`, `type`) attached:

```{r}
#| label: as-dataframe
#| eval: !expr has_sf
df <- as.data.frame(dk())
names(df)
```

## Checking and converting

`is_ggseg_atlas()` tests whether an object has the right class:

```{r}
#| label: is-ggseg-atlas
is_ggseg_atlas(dk())
is_ggseg_atlas(mtcars)
```

`as_ggseg_atlas()` coerces lists with the right structure into a proper
`ggseg_atlas`:

```{r}
#| label: as-ggseg-atlas
atlas_list <- as.list(dk())
recovered <- as_ggseg_atlas(atlas_list)
is_ggseg_atlas(recovered)
```

If you have an atlas object from an older version of ggseg that stored sf data
directly in `$data` instead of using the new `ggseg_atlas_data` wrapper,
`convert_legacy_brain_atlas()` will migrate it to the unified format.
