---
title: "Migrating atlases and code"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Migrating atlases and code}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup-opts
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r}
#| label: setup
library(ggseg.formats)
```

This vignette covers the two migrations ggseg.formats has asked of its users:
re-keying `region` in your own data tables, and rewriting an atlas package's
geometry into the sf-optional format.

## Migrating code to the 0.1.0 region keys

Before 0.1.0 the `region` column of the bundled atlases held a long-form display
name (`"banks of superior temporal sulcus"`). From 0.1.0 it holds a short,
hemisphere-free key derived from `label` (`"bankssts"`), and the long-form name
lives in the new `names` column. `label` never changed and remains the stable
join key.

Nothing errors when this bites. `merge(all.x = TRUE)` and the ggseg fill scale
both yield `NA` for an unmatched region, which draws as a blank parcel — so the
only symptom is a figure with holes in it.

`legacy_region_map()` translates a table of old values mechanically:

```{r}
#| label: legacy-map
#| eval: false
map <- legacy_region_map(dk())
my_data$region <- unname(map[my_data$region])
```

Every pre-0.1.0 `region` value of all four bundled atlases translates: `dk`
35/35, `aseg` 19/19, `tracula` 26/26 and `suit` 13/13, whose keys never changed,
so its map is the identity. `aseg`'s `"Thalamus Proper"` translates too, to
`"thalamus"`, even though the alias row it lived on is gone from `core`. For any
other atlas the map is zero-length, because only the bundled four were re-keyed.

### Do not use `names` for this

`names` is the curated display name, spelled to read well in a legend, and for
several regions it deliberately differs from what 0.0.4 shipped as `region`:
`aseg` has `names` `"ventral diencephalon"` where 0.0.4 had `region`
`"ventraldc"`, and `"corpus callosum posterior"` where 0.0.4 had
`"cc posterior"`. `legacy_region_map()` reads a recorded table of the 0.0.4
values instead, so it stays correct however `names` is later improved.

If a pattern you pass to `atlas_region_remove()`, `atlas_region_keep()` or
`atlas_region_contextual()` matches nothing, they now warn and, where the
pattern looks like a pre-0.1.0 name, say which key to use instead.

### The accessors return rows, not a set

`atlas_regions()`, `atlas_labels()` and `atlas_names()` used to return a sorted,
de-duplicated, `NA`-dropped set, so the three disagreed on length for the same
atlas and could not be zipped. Each now returns its `core` column unchanged: one
element per `core` row, in row order, repeats and `NA`s retained. So
`atlas_labels(a)[i]`, `atlas_regions(a)[i]` and `atlas_names(a)[i]` describe the
same row.

In practice `atlas_regions()` and `atlas_names()` now repeat values, once per
hemisphere or view row, and none of the three is sorted any more. Code that
indexed the old output positionally, or relied on it being a set, wants
`unique()` — and `sort(unique(x))` reproduces the old value exactly.
`atlas_views()` is unchanged: views are a separate axis, not a `core` column.

## Migrating an atlas package to sf-optional

Since the **sf-optional** milestone (ggseg 2.2), a `ggseg_atlas` stores its 2D
geometry in a single `geom` slot that can hold either an `sf` table or the
sf-optional `brain_polygons` representation. The polygon form renders
identically through the `geom_polygon`-based path in ggseg, but carries no
dependency on the sf class machinery — so atlases keep working on wasm builds
and air-gapped installs where sf (and its GDAL/GEOS/PROJ system libraries)
cannot be installed.

This vignette is for maintainers of downstream atlas packages
(`ggsegFreeSurfer`, `ggsegSchaefer`, `ggsegGlasser`, …). Your package ships
`brain_atlas`/`ggseg_atlas` objects as `.rda` files under `data/`. Migrating
means rewriting those files once so the geometry is stored as polygons, then
dropping sf from your `DESCRIPTION`.

### The recipe

From the root of your atlas package, run:

```r
# 1. rewrite every atlas in data/ into the polygon format
ggseg.formats::migrate_atlas_files("data")

# 2. drop sf from DESCRIPTION (it is no longer needed at install or run time)
usethis::use_package("sf", type = "Suggests")

# 3. rebuild the package data documentation and reinstall
devtools::document()
```

That is the whole migration. Commit the rewritten `data/*.rda`, push, and
release.

### What `migrate_atlas_files()` does

It walks the directory, loads each `.rda`, finds every atlas object inside,
converts its `geom` to `brain_polygons`, drops any legacy `sf`/`polygons`
slots, and saves the file back with `xz` compression. Files with nothing to
migrate are left untouched, and it reports what it changed:

```r
ggseg.formats::migrate_atlas_files("data")
#> ✔ Migrated `dk.rda`.
#> ✔ Migrated `aseg.rda`.
#> ℹ Skipped `palette.rda` (nothing to migrate).
```

The conversion reads sf coordinates, so **sf must be installed on the machine
you run the migration from**. This is a one-time maintainer step; the published
package no longer needs sf.

It is idempotent — running it twice is a no-op on already-migrated files, so it
is safe to wire into a `data-raw/` build script.

#### Keeping sf instead

If your atlas package genuinely needs sf geometry (for example it exposes
geometric operations), pass `keep_sf = TRUE` to normalise everything into the
single `geom` slot as sf rather than polygons:

```r
ggseg.formats::migrate_atlas_files("data", keep_sf = TRUE)
```

### Verifying the result

After migrating, the geometry is sf-optional and round-trips losslessly. You
can rehydrate sf on demand with `as_sf_atlas()`, and go back with
`as_polygon_atlas()`:

```{r}
#| label: roundtrip
poly <- as_polygon_atlas(dk())
is_atlas_polygon(poly)

atlas_labels(poly) |>
  head()
```

A migrated atlas plots through ggseg with no sf installed. If a lite-only
install meets an atlas that is *still* sf-backed, `as_polygon_atlas()` converts
it on the fly when sf is available, and otherwise aborts with a message naming
`migrate_atlas_files()` — the signal that the atlas package itself needs the
one-time migration above.
